---
title: "Notebook 01: Download Raw Data"
author: "Shirshak Aryal"
date: today
format:
  html:
    code-overflow: scroll
    embed-resources: true
    theme:
      light: flatly
      dark: [darkly, ../custom_darktheme.scss]
    respect-user-color-scheme: true
execute:
  echo: true
  message: false
  warning: false
---

Download **raw RNA-seq counts** and **clinical metadata** for **TCGA-GBM** from the **NIH Genomic Data Commons** (the 'GDC' in GDCQuery) using the `TCGAbiolinks` library.

## 1. Import Libraries

```{r}
#| label: import_libraries


library(TCGAbiolinks)  # query and download TCGA datasets from GDC
library(tidyverse)     # data manipulation and visualisation

cat("\nLibraries loaded successfully\n")
cat("TCGAbiolinks version:", as.character(packageVersion("TCGAbiolinks")), "\n")
cat("tidyverse version:", as.character(packageVersion("tidyverse")), "\n")
```

## 2. Download Dataset

### 2.1. Query For Dataset Details Before Downloading

The **TCGA-GBM (The Cancer Genome Atlas - GlioBlastoma Multiforme) dataset** is a **multi-omic** dataset, meaning it has multiple **modalities** of data from the same collection of samples (e.g. bulk RNA-seq, DNA-seq, etc.). The focus of this project is only the **bulk RNA-seq modality.**

In the query below:

-   The **`data.category`** **parameter** being **"Transcriptome Profiling"** specifically, together with the **`workflow.type`** **parameter**, denotes that the data queried for is of the **bulk RNA-seq** modality.
-   The **`data.type`** **parameter** being **"Gene Expression Quantification"** specifically, denotes that gene-level counts are retrieved, as opposed to [transcript-level reads or the raw sequencing files]{.underline}.
-   The **`workflow.type` parameter** being **"STAR - Counts"** denotes that the version of the retrieved data is specifically the one that was **both aligned AND counted by the STAR software,** as opposed to the [historical]{.underline} version where [STAR was only used for alignment]{.underline} and the counting was done by a separate tool called [HTSeq-count]{.underline}**.**

```{r}
#| label: query_for_dataset_details

query_rnaseq <- GDCquery(
    project = "TCGA-GBM",                          # i.e. the TCGA-GBM dataset
    data.category = "Transcriptome Profiling",     # i.e. bulk RNA-seq modality (combined with workflow.type below)
    data.type = "Gene Expression Quantification",  # i.e. raw gene counts specifically
    workflow.type = "STAR - Counts"                # i.e. latest version; together with data.category, filters to bulk RNA-seq
)

results <- getResults(query_rnaseq)
cat("\nQuery executed successfully")
```

### 2.2. Inspect Query Results

The **`results`** **variable** contains the query results, which is simply a dataframe containing **metadata about the actual bulk RNA-seq data.**

-   **Rows** represent **each sample involved** in the experiment that generated the bulk RNA-seq data.

-   **Columns** represent **details about each of those samples.**

#### 2.2.1. Inspect Results Structure

```{r}
#| label: inspect_results_structure


cat("\nQuery results (first 5 columns shown horizontally for visual clarity):\n")
results[, 1:5] %>% glimpse()

# Polished tabular view of key columns
results %>%
  select(cases, sample_type, data_format, access, file_size) %>%
  head(10) %>%
  knitr::kable(caption = "First 10 rows of query results (selected metadata columns)")
```

#### 2.2.2. Visualize Sample Distribution

```{r}
#| label: visualize_sample_distribution


# visualize sample type distribution as a table

results %>%
  count(sample_type) %>%
  rename("Sample Type" = sample_type, "Count" = n) %>%
  knitr::kable(caption = "TCGA-GBM samples by biological condition")
```

### 2.3. Filter And Download Datasets Of Samples

The 14 recurrent tumor and 5 normal samples will be **entirely excluded**, focusing only on the **372 Primary Tumor samples** for comparison of **Proneural vs Mesenchymal** subtypes.

The download will save the **raw sample files** (which are TSV) to the **`01_data/01_raw`** **directory**.

#### 2.3.1. Filter To Primary Tumor Samples

```{r}
#| label: filter_to_primary_tumor_samples


query_primary <- GDCquery(
  project = "TCGA-GBM",
  data.category = "Transcriptome Profiling",
  data.type = "Gene Expression Quantification",
  workflow.type = "STAR - Counts",
  sample.type = "Primary Tumor"  # new parameter added to the previous query
)

results_primary <- getResults(query_primary)
cat("Primary tumor samples:", nrow(results_primary), "\n")
```

#### 2.3.2. Download Filtered Dataset

```{r}
#| label: download_filtered_dataset


GDCdownload(
  query = query_primary,
  directory = "../01_data/01_raw/01_samples",
  files.per.chunk = 50
)

# Confirm download
# list.files() returns a list of filenames (as strings)
downloaded_files <- list.files(
  path = "../01_data/01_raw/01_samples",  # the directory to look inside for the filenames
  recursive = TRUE,                       # look inside subfolders too
  pattern = "\\.tsv$"                     # take only those filenames that end in .tsv
)

cat("Download complete\n")
cat("TSV files found:", length(downloaded_files), "\n")
cat("Expected:", nrow(results_primary), "\n")
cat("Match:", length(downloaded_files) == nrow(results_primary), "\n")
```

## 3. Download Clinical Metadata

### 3.1. Query For Clinical Metadata

```{r}
#| label: query_for_clinical_metadata


clinical <- GDCquery_clinic(
  project = "TCGA-GBM",
  type = "clinical"
)
cat("\nQuery executed successfully")
```

### 3.2. Inspect Clinical Metadata

```{r}
#| label: inspect_clinical_metadata


cat("\nClinical metadata (first 5 columns shown horizontally for visual clarity):\n")
clinical[, 1:5] %>% glimpse()

clinical %>%
  select(submitter_id, vital_status, days_to_death,
         days_to_last_follow_up, age_at_index) %>%
  head(10) %>%
  knitr::kable(caption = "Key clinical variables (first 10 patients)")
```

### 3.3. Save Clinical Metadata

```{r}
#| label: save_clinical_metadata


saveRDS(clinical, "../01_data/01_raw/02_clinical/01_clinical_metadata.rds")
cat("Saved to 01_data/01_raw/02_clinical/01_clinical_metadata.rds\n")
```

## 4. Perform Sanity Checks

```{r}
#| label: perform_sanity_checks


# Confirm sample files downloaded
sample_files_exist <- length(list.files(
  "../01_data/01_raw/01_samples",
  recursive = TRUE,
  pattern = "\\.tsv$"
)) == 372

# Confirm clinical metadata saved
clinical_file_exists <- file.exists(
  "../01_data/01_raw/02_clinical/01_clinical_metadata.rds"
)

cat("Sample files (372 TSV):", ifelse(sample_files_exist, "✓", "✗"), "\n")
cat("Clinical metadata:     ", ifelse(clinical_file_exists, "✓", "✗"), "\n")

if (sample_files_exist && clinical_file_exists) {
  cat("\nNotebook 01 complete — all outputs verified\n")
} else {
  cat("\nWARNING: One or more outputs missing\n")
}
```
